Active Learning: Because Guesswork is Overrated

### 2025-02-03 doc/paper.tex
Add finished report

\twocolumn[
\icmltitle{TODO: Funky active learning pun}

@@

\begin{icmlauthorlist}
	\icmlauthor{Lukas Weber}{first}
\end{icmlauthorlist}

@@

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
	A lot of unlabelled data generally available but annotation is costly. A goal in fields where labelled data is not available  is to maximize performance with minimum amount of labeled data instances. \\
	Active learning selects instances for labelling based on a selection strategy. 
	\begin{itemize}
		\item Objective: Compare active learning query methods using modAL.
		\item Methods: Experiments with Logistic Regression (MNIST) and a CNN (CIFAR).
		\item Experiments: Vary initial dataset sizes (small, moderate, large).
		\item Key findings: Highlight main conclusions about performance in different scenarios.
	\end{itemize}
\end{abstract}

@@

Supervised learning often requires large, annotated datasets, but labeling data can be costly and time-intensive, especially when expert knowledge is needed. 
Active learning addresses this challenge of reducing the overall labeling costs through the selection of the most informative samples for annotation based on a query strategy. This approach enables machine learning models to perform well in tasks with limited data and high labeling costs. \\
The effectiveness of active learning depends heavily on the chosen query strategy, which determines which samples are selected for annotation. Two main frameworks are commonly used:
\begin{enumerate}
	\item Pool-based active learning: A model is trained on a small labeled dataset and used to evaluate a large pool of unlabeled data. The most informative samples are selected based on a query strategy, labeled by an oracle, and added to the training set. The model is retrained with the updated training set and the process is repeated until a stopping criterion is met.
	\item Stream-based active learning: Here, unlabeled data arrives sequentially and the model decides whether to query the label of an incoming instance using an evaluation metric like the uncertainty of its prediction and checking whether a certain threshold is reached. If not, the instance is discarded.
\end{enumerate}
This report focuses on comparing several query strategies for both pool-based and stream-based active learning. Experiments in this paper leverage the modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning}, which simplified the implementation of active learning pipelines, includes multiple pre-implemented query strategies and is compatible with models from the Scikit-learn library~\cite{scikit-learn}. \textcolor{red}{CNN implemented with PyTorch(ADD CITATION) and transformed with skorch~\cite{skorch} to make it compatible with modAL as it is only compatible with sklearn models.} \\
Several previous works have explored and compared different query methods across various domains. \\
\cite{schröder_revisitinguncertaintybasedquerystrategies} compare various uncertainty-based query strategies in the context of fine-tuning transformer models in text classification tasks. \cite{zhan_comparativesurveydeepactive} on the other hand compares several query strategies on the image datasets MNIST and CIFAR. \\

@@

A recurring issue in active learning research is the lack of standardization in experimental setups, as noticed additionally by~\cite{werner_comparableactivelearning}. This inconsistency complicates the comparability of existing and novel query strategies. To address this, they propose a benchmark framework for evaluatting active learning strategies across tabular, image and text datasets. This framework uses robust evaluation protocols to reduce variance and ensure comparability. \\
In contrast, \cite{ueno_benchmarkingofquerystrategies} focus specifically on image-based datasets from diverse domains such as medical or visual inspection images. While~\cite{werner_comparableactivelearning} focuses on a wide range of data-domains including images, text and vector-based datasets, this report focuses on comparing strategies in the image domain. 
It compares query strategies for both models such as simple logistic regression on MNIST and more complex architectures, like a CNN on CIFAR-10 to study the effect of model complexity on the optimal query strategy. \\
It is uncertain whether well working query strategies generalize well from simple to more complex models like Neural Networks~\cite{schröder_surveyactivelearningtext}. Understanding this generalizability is crucial for real-world applications where model complexity varies across tasks and datasets. \\
Furthermore, this report investigates the impact of the initial labeled dataset size, as its impact in combination with model complexity is rarely explored. 
By examining scenarios with varying amounts of initial labeled data, this work simulates various real-world applications: (1) starting with no labeled data, (2) having a small pool of labeled data, and (3) working with a well-trained model to identify data points for fine-tuning.

@@

	\item \textbf{Systematic evaluation of active learning query strategies} across different datasets and model complexities, using MNIST with logistic regression and CIFAR-10 with a CNN.
	\item \textbf{Investigation of the incluence of model complexity} on the effectiveness of query strategies to provide insights into how shallow and deep models respond differently to active learning methods.
	\item \textbf{Analysis of impact of initial labeled dataset size} on the performance of query strategies, simulating possible real-world scenarios such as starting from scratch, having a small labeled dataset or fine-tuning a pre-trained model.

@@

Section~\ref{sec:methods} outlines the used datasets, processing steps and models. Section~\ref{sec:experiments} describes the experimental design, followed by the results and discussion in sections~\ref{sec:results} and~\ref{sec:discussion} respectively. The report concludes with key findings in section~\ref{sec:conclusion}

\section{Methods}\label{sec:methods}
Several different datasets are used in the following experiments.
\subsection{Datasets}
MNIST is a dataset consisting of handwritten digits with 60.000 train samples and 10.000 test samples. Digits are size-normalized and centered in a $28 \times 28$ image~\cite{mnist}. Each digits belongs to one of 10 classes, corresponding from 0 to 9. \\
Upon loading the dataset, each image is vectorized, i.e. transformed from a $28 \times 28$ sized image-matrix to a vector of length $784$. Furthermore, each pixel contains a grayscale-value from 0 to 255. Pixel values are normalized to a range between 0 and 1 to provide a value range that works well with logistic regression. 
\\
\textcolor{red}{Here describe the validation split}
\\
\\
CIFAR-10 is a dataset which consists of 50.000 train- and 10.000 test-images. Each natural image belongs to one of 10 classes. Each class has exactly 5.000 train and 1.000 test images. The image-size is $32 \times 32$ pixels. 
\textcolor{red}{CIFAR preprocessing split and validation split}
\\
\\
Both used datasets are balanced, i.e. each class has roughly the same number of samples as any other class.

\subsection{Models}
\begin{itemize}
	\item Logistic Regression: Details on implementation and hyperparameters, how were they chosen?
	\item same for CNN: CrossEntropyLoss (combines Log Softmax and NLLoss in one single class)
\end{itemize}
Are models re-initialized after each new data points and newly trained or fine-tuned?

\subsection{Active Learning}
\begin{itemize}
	\item How are the splits (initial pool, initial training data) generated
	\item numbmer of iterations for active learning
	\item Used active learning query strategies
\end{itemize}

Metrics for evaluation (Accuracy, train error, confusion matrix, ...)

\section{Experiments}\label{sec:experiments}
Setup: details for reproducibility (random number generators)?
Random seed = 42
\cite{comparableactivelearning} mention reproducibility problems in models even when setting random number generators before each run. By training models multiple times in succession and checking whether the same configurations lead to different outcomes, it is shown that in this scenario, reproducibility is guaranteed. \textcolor{red}{TODO: Schnell ein skript aufsetzen und dann finale accuracy und so weiter vergleichen.}
\\
Did I perform multiple runs?
\\

Scenarios
\begin{itemize}
	\item describe experiments and their settings and purposes
\end{itemize}

\section{Results}\label{sec:results}
Quantitative results: Tables or plots comparing the query strategies across scenarios. Also maybe key perforamnce differences (accuracy vs number of labeled examples)

\section{Discussion}\label{sec:discussion}
Qualitative analysis: Why certain strategies perform better or worse in specific settings. Trends between simple models and complex models
\\
\\
Impact of initial dataset size
\\
\\
Most methods require knowledge about the balance of the dataset and might be disadvantageous to use in case of a disbalanced dataset. In real world scenarios, this is not known. Both used datasets are balanced so the impact of query strategies on imbalanced datasets was not tested. \\
Only image-based datasets were tested. Not vector- or text-based.

\section{Conclusion}\label{sec:conclusion}
Summary of findings and insights (maybe that a strategy works best with small datasets while another one is robust on bigger ones)
\\
Implications: practical recommendations?
\\
Future work? (More diverse datasets, exploring additional strategies, applying in real-world tasks)

\newpage

\section{Cool ideas}
\begin{itemize}
	\item Show the typical framework of active learning in a figure in the introduction
	\item Accuracy vs. fraction of data size in relation to the datset (percentage of used data points)
\end{itemize}

\section{Good formulations}
From the active learning survey: Entropy is an information-theoretic
measure that represents the amount of information needed to “encode” a distri-
bution. As such, it is often thought of as a measure of uncertainty or impurity in machine learning.

\section{Comparison to the other papers}
% FROM HERE THIS IS NOT GOING TO BE PART OF THE FINAL PART
Previous papers do not: Is everything here now mentioned previously? if yes, then it can be deleted.
\begin{itemize}
	\item \cite{schröder_revisitinguncertaintybasedquerystrategies}: Does not compare across different sizes of the pool, and for simple models.
	\item \cite{comparativesurveydeepactive}: Does not compare impact of different initial train set size and strategies on shallow classifiers.

	\item According to \cite{ueno_benchmarkingofquerystrategies}: \textcolor{red}{check if true:} Does not compare different sizes of the initial starting sets as well as how simple and more complex perform with the query methods

	\item \cite{werner_comparableactivelearning}: They focus on a wide range of domains, this paper only compares methods in the image domain. I compare strategies for a simple logistic regression model and a complex CNN across different starting sizes of the annotated dataset. \\
\end{itemize}
% TO HERE

>>>>

\twocolumn[
\icmltitle{Active Learning: Because Guesswork is Overrated}

@@

\begin{icmlauthorlist}
	\icmlauthor{Lukas Weber, lukas.weber@studenti.unitn.it}{first}
\end{icmlauthorlist}

@@

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
	A huge amount of unlabeled data is typically available but annotating large datasets is often costly and time-consuming, making active learning a crucial tool for selecting the most informative samples to minimize annotation efforts. This report compares different active learning query strategies using the modAL framework in a pool-based setting. In experiments, a logistic regression model and an ensemble of such models is evaluated on the MNIST dataset. The impact of different query strategies is analyzed, including Uncertainty Sampling, Margin Sampling, Entropy Sampling (including ensemble adaptions) and Ranked batch-mode Sampling. Findings indicate that querying samples with Margin Sampling achieves the highest accuracy and fastest convergence. While ensemble models show faster convergence with small datasets, they provide shrinking improvements over single models as dataset size increases. These results highlight the high importance of query strategy selection on model performance in active learning pipelines and explore conditions under which ensemble models may improve performance.
\end{abstract}

@@

Supervised learning often requires large, annotated datasets. However obtaining high-quality labeled data can be costly and time-intensive, especially expert knowledge is required, such as medical imaging. Though, unlabeled data is usually easily available. 
Active learning addresses this challenge by reducing the overall labeling costs through the selection of most informative samples for annotation based on a query strategy. This approach enables machine learning models to perform well in tasks with limited data and high labeling costs. \\
The effectiveness of active learning is tied to the chosen query strategy, that determines which unlabeled samples are selected for annotation and then added to the training set. \\
This report evaluates several strategies in a pool-based active learning setting: A model is initially trained on a small labeled dataset and used to evaluate a large pool of unlabeled data. The most informative samples are selected based on a query strategy, they are then labeled by an oracle, and added to the training set. The model is retrained with the updated training set and the process is repeated until a stopping criterion is met.
Experiments in this paper leverage the modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning}, which simplifies the implementation of active learning pipelines, includes multiple pre-implemented query strategies and is compatible with Scikit-learn models~\cite{scikit-learn}. \\
The experiments focus on the image domain, using the MNIST dataset~\cite{mnist}, and examine the impact of query strategies on a single logistic regression model and a committee-based approach.
Several previous works have explored and compared different query methods across various domains. \cite{schröder_revisitinguncertaintybasedquerystrategies} compared various uncertainty-based query strategies in the context of fine-tuning transformer models in text classification tasks, while \cite{zhan_comparativesurveydeepactive} examined several query methods in image classification tasks using MNIST and CIFAR. \\

@@

A recurring issue in active learning research is the lack of standardization in experimental setups, making it difficult to compare existing and new query strategies, as noticed additionally by~\cite{werner_comparableactivelearning}. This inconsistency complicates the comparability of existing and novel query strategies. To address this, they propose a benchmark framework for evaluating active learning strategies across tabular, image and text datasets. This framework uses robust evaluation protocols to reduce variance and ensure comparability. \\
In contrast, \cite{ueno_benchmarkingofquerystrategies} focus specifically on image-based datasets from diverse domains such as medical or visual inspection images. While~\cite{werner_comparableactivelearning} focuses on a wide range of data-domains including images, text and vector-based datasets.
This work focuses on query strategy comparisons within the image domain, specifically for a logistic regression model and an ensemble-based approach on MNIST. \\

@@

	\item A systematic evaluation of active learning query strategies using \textbf{logistic regression on MNIST}.
	\item \textbf{Investigation of the incluence of model aggregation} on the effectiveness of query strategies to provide insights into how these models respond differently to active learning methods.

@@

Section~\ref{sec:methods} outlines the used dataset, processing steps and models and describes the experimental design, followed by the results and discussion in sections~\ref{sec:results} and~\ref{sec:discussion} respectively. The report concludes with key findings in section~\ref{sec:conclusion}. \\
The corresponding implementation can be found on \textcolor{red}{\href{https://github.com/LukaslWeber/NLP_II_Interactive_Learning}{Github}}.

\section{Methods}\label{sec:methods}
The following section describes the dataset which was used in experiments, preprocessing steps, the logistic regression model and other parameters which are needed for reproducibility.
\subsection{Dataset}
MNIST is a dataset consisting of handwritten digits with 60.000 train samples and 10.000 test samples. Digits are size-normalized and centered in a $28 \times 28$ image~\cite{mnist}. Each digit belongs to one of 10 classes, corresponding to the  numbers from 0 to 9. \\
Upon loading the dataset, each image is vectorized, i.e. transformed from a $28 \times 28$ sized image-matrix to a vector of length $784$. Furthermore, each pixel initially contains a grayscale value ranging from 0 to 255. Pixel values are additionally normalized to a range between 0 and 1 in order to provide a value range that works well with logistic regression.
%\\
%\\
%CIFAR-10 is a dataset which consists of 50.000 train- and 10.000 test-images. Each natural image belongs to one of 10 classes. Each class has exactly 5.000 train and 1.000 test images. The image-size is $32 \times 32$ pixels. 
%\textcolor{red}{CIFAR preprocessing split}. CIFAR-10 includes real-world images and is thus expected to be more complicated to learn.
%
MNIST is a class-balanced dataset, i.e. each class has roughly the same number of samples as any other class.

\subsection{Multinomial Logistic Regression Model}
Sklearn's LogisticRegression module is used to predict digits with the given data from MNIST. Multinomial logistic regression predicts a probability for each possible class instead of a single probability as in Binomial logistic regression.
\\
The model estimates the probability $P(y=i \mid X)$ of a sample belonging to class $i$ using the SoftMax function:
\begin{align}
	P(y=i \mid X)=\frac{\exp \left(X \beta_i\right)}{\sum_{j=1}^k \exp \left(X \beta_j\right)}
\end{align}
where $\beta_i$ are the coefficients for class $i$. The SoftMax function ensures that probabilities across classes sum to 1 and consequently that a model creates a probability distribution over the classes. \\
During training, the model maximizes the likelihood of the observed data:
\begin{align}
	\mathcal{L}(\beta)=\prod_{n=1}^N \prod_{i=1}^k\left(P\left(y_n=i \mid X_n\right)\right)^{y_{n i}}
\end{align}
where $N$ is the number of samples, and $y_{n i}$ indicates if sample $n$ belongs to class $i$. $k=10$ denotes the number of classes. \\
The likelihood of the observed data is maximized by minimizing the Cross-Entropy loss $\mathcal{L}$ which is defined as:
\begin{align}
	\mathcal{L}=-\frac{1}{N} \sum_{n=1}^N \sum_{j=1}^k y_{n j} \log \left[P\left(y=j \mid X_n\right)\right]
\end{align}
To find hyperparameters which maximize performance on the given dataset, 5-fold-cross-validation is used with a grid-search, testing out a variety solvers, regularization methods and corresponding strengths, maximum iterations and stopping-tolerance options. \\
The resulting models use the "newton-cg" solver which combines Newton's method with conjugate gradient descent. It uses the gradient and the second order derivative of $\mathcal{L}$ given by the Hessian matrix to iteratively find the minimum of $\mathcal{L}$. \\
Regularization using the $\ell_2$-norm prevents overfitting with a strength $C=0.01$ in addition to a stopping-criterion $tol=0.0001$, which stops the learning process when the $\ell_2$-norm of the gradient of $\mathcal{L}$ is lower than $tol$, indicating convergence. \\
For reproducibility purposes, the same random seed ($42$) is used to initialize all random number generators and single model weights. Weights of models in an ensemble are initialized with a fixed set of random seeds (42, 43, 44, 45, 46) to introduce diversity. These measures ensure that same configurations lead to the same outcomes, even when training models in succession.

\subsection{Active Learning}
In the following experiments, a single logistic regression model is compared to a committee of 5 models across different query strategies. The committee predicts by averaging the class probabilities obtained by each learner, called the consensus probabilities. The final prediction is the class with the highest consensus probability. \\
In pool-based active learning, an initial labeled dataset is required. When a single model is used for training and querying new samples, $n_{init}=75$ labeled instances are used to train the model initially, ensuring a moderate fit before querying samples. This is done to make sure that initial samples are chosen on a meaningful basis and do not bias the rest of the training negatively. For the committee, each model is initialized with $60\%$ randomly chosen samples from the initial labeled train set, resulting in $45$ initial data points per model. \\
For both methods, models are trained for $init\_max\_iter=50$ epochs on the initial dataset. $init\_max\_iter$ is based on the hyperparameters found by 5-fold-cross-validation and ensures a sufficient fit to the datapoints. Thus, reducing the bias from early queried samples which were chosen when model predictions are inaccurate. \\
After initial training, 5 samples are queried at a time for 1000 iterations, resulting in 5000 additional samples at the end of the learning process. After quering, the new samples are added to the training set and models are fine-tuned for $n_{iter}=20$ iterations. 
\subsection{Query Strategies}
A multitude of query strategies are used to compare a single model to an ensemble of models. As a control metric, Random Sampling is used to give a baseline for other query strategies.\\
Uncertainty Sampling selects the least certain instances for labeling.
\begin{align}
	Uncertainty(\mathbf{x})=1-P_\theta(\hat{y} \mid \mathbf{x})
\end{align}
Where $\hat{y}=\operatorname{arg}\underset{y}{\operatorname{max}} \ p_\theta(y \mid \mathbf{x})$, i.e. corresponds to the predicted class of the model and $P_\theta$ is the probability to predict class $\hat{y}$ with the model with parameters $\theta$. \\
Entropy Sampling selects instances with the highest entropy to query for labeling, focusing on the most ambiguous cases. It measures the amount of information needed to encode a distribution and is often used as a measure of uncertainty.
\begin{align}
	\mathcal{H}(\mathbf{x})=-\sum_{i=0}^{9} P_\theta(y_i \mid \mathbf{x}) \log P_\theta(y_i \mid \mathbf{x})
\end{align}
Where $y_i$ stands for the probability to predict class $i$. \\
Vote Entropy Sampling is a query strategy designed for ensemble models. It calculates the ensemble predictions, then calculates the probability distribution over the predictions which are used to determine the Entropy $\mathcal{H}$. Vote Entropy then selects instances with the highest Entropy. \\
Margin Sampling selects the instances where the difference between the most likely and second most likely class is the smallest, indicating high ambiguity.
\begin{align}
	Margin(\mathbf{x})=P_\theta\left(\hat{y}_1 \mid \mathbf{x}\right)-P_\theta\left(\hat{y}_2 \mid \mathbf{x}\right)
\end{align}
where $\hat{y}_1$ and $\hat{y}_2$ are the most likely and the second likely predicted labels given by $\hat{y}_1=\operatorname{arg}\underset{y}{\operatorname{max}} \  p_\theta(y \mid x)$ and $\hat{y}_2=\operatorname{arg}\underset{y \neq \hat{y}_1}{\operatorname{max}} \  p_\theta(y \mid x)$, respectively. \\
Ranked batch-mode Sampling~\cite{ranked_batch_mode_query_strategy} weights uncertainty and diversity. It is designed for querying multiple samples, as traditional approaches such as uncertainty sampling may add redundant samples when querying multiple samples at once. \\
\begin{align}
	RBM(x)&=\alpha \left(1-\Phi\left(x, X_{\text {labeled }}\right)\right) \notag \\ &+(1-\alpha) \ Uncertainty(x)
\end{align}
Where $\alpha=\frac{\left|X_{\text {unlabeled}}\right|}{\left|X_{\text {unlabeled}}\right|+\left|X_{\text {labeled}}\right|}$, $X_{\text {labeled}}$ and $X_{\text {unlabeled}}$ are the labeled dataset and unlabeled pool, respectively. $\Phi$ denotes the similarity given by the Cosine Similarity metric.
\begin{figure*}[!h]
	\begin{center}
		\includegraphics[width=0.95\linewidth]{fig/test_accuracy.pdf}
	\end{center}
	\caption{Test accuracy versus the number of annotated samples that were added to the train dataset by querying instances. As examples are queried and added to the training dataset, the model is retrained, and its performance is evaluated on the test set.}
	\label{accuracy_plot}
\end{figure*}
\begin{figure*}[!h]
	\begin{center}
		\includegraphics[width=0.95\linewidth]{fig/conf_mat_margin_sampling.pdf}
	\end{center}
	\caption{Confusion Matrix of models trained with Margin Sampling.}
	\label{conf_mat_margin_sampling}
\end{figure*}
\subsection{Evaluation metrics}
In the following section, multiple evaluation metrics are used to gain insights into understanding the differences and quality of each query method and the single-model vs. ensemble-model approach. \\
Due to the fact, that MNIST is class-balanced, a well-working metric is the accuracy. It is defined as as the number of correctly classified instances divided by the total number of instances. \\
Furthermore, confusion matrices are created to evaluate systemic errors in the prediction of certain numbers.  
\section{Results}\label{sec:results}
A Logistic Regression model that is trained on the whole labeled train dataset of MNIST reaches a test accuracy of $92.58\%$. This serves as a theoretical maximum that can be reached given a sufficient amount of train data. \\
By plotting the accuracy on the test samples vs. the number of queried examples in figure~\ref{accuracy_plot}, convergence behaviour of different query methods can be analyzed. On MNIST, Margin Sampling reaches the highest Accuracy with 92.08\% for both models and converges the fastest, followed by Random Sampling. Entropy Sampling and Uncertainty Sampling perform worse than Random Sampling. Only after querying 5000 additional samples, the difference in accuracy between both strategies and Random Sampling is 0.25\%. For most query strategies, performance between a single model and an ensemble of models does not differ. \\
Furthermore, it can be seen that ensemble models converge faster than single models with the same query method, with small train datasets, but differences between models shrink with an increased number of queried samples. \\
The ensemble that is trained with Vote Entropy sampling performs better than the models trained with Entropy sampling with an accuracy difference of 2.08\% to the single model and 1.79\% to the ensemble, indicating that query strategies specifically adapted for ensemble models can boost classification performance. \\
Figure~\ref{conf_mat_margin_sampling} and shows the confusion matrices of the best performing query strategy, Margin Sampling. The single model trained with Margin Sampling achieves the lowest accuracy of $86\%$ when an input image contains the number 5, whereas the ensemble reaches has 1\% more accuracy in this case, which also resembles the worst number for the ensemble. In both cases, the most likely prediction is 3 when the number 5 is shown.

\section{Discussion}\label{sec:discussion}
Results show that Margin Sampling achieves the highest accuracy and fastest convergence by prioritizing instances with the greatest ambiguity, effectively adding diversity to the training dataset. It does not show high systemic errors. In contrast, Uncertainty and Entropy Sampling underperform compared to Random Sampling, possibly due to redundant sample selection. \\
Across query strategies, the ensemble converges faster than single models the smaller train datasets are, leveraging prediction diversity in early stages. However, its advantage diminishes as the labeled train dataset grows. This leads to the conclusion that ensembles help to reduce the impact of single model biases in small datasets. For most strategies, the ensemble's added diversity does not translate into substantial performance gains, as shown by its marginal improvement over single models. Both, single and ensemble models, converge similarly for the same sampling strategy, indicating that a single Logistic Regression Model is able to effectively learn MNIST when given enough data. Furthermore, early biases can be reduced by using an ensemble. Using more complex models and data might lead to the ensemble converging to a higher accuracy than single models, as is often seen. \\
This study uses a logistic regression model and a simple, balanced dataset, which simplifies the analysis but limits generalizability. While the findings highlight the effectiveness of certain strategies for shallow models, it is unclear if these results hold for more complex models or imbalanced datasets.
Future work should explore diverse data types, like text, tabular or time series data, complex models, and imbalanced datasets to better understand the broader applicability of these findings and see whether complex models show similar convergence profiles.

\section{Conclusion}\label{sec:conclusion}
This report systematically evaluated active learning query strategies using logistic regression on MNIST. Margin Sampling is shown to be the most effective strategy, achieving the highest accuracy and fastest convergence by focusing on instances with the greatest ambiguity. Ensemble models demonstrated faster convergence in early training stages, possibly due to higher diversity but offered limited performance gains over single models as labeled datasets grew. \\
Key findings suggest that ensembles are particularly advantageous when dealing with small datasets, reducing biases from single-model predictions. However, since single logistic regression models already perform well on MNIST when there’s enough labeled data, using an ensemble doesn’t add much benefit in this case. The underperformance of uncertainty- and entropy-based strategies compared to Random Sampling indicates that redundant sample selection remains a challenge, as Random Sampling may unintentionally add variety to the data. \\
Future research should extend these insights to more complex models and datasets, including imbalanced and multi-domain scenarios such as text and tabular data. Additionally, exploring novel query strategies tailored to diverse use cases could further optimize active learning pipelines for real-world applications.
